\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-20 paper.qmd
Add draft for simulated data

## Import data
df_2 <- read_rds("data/five_gau_clusters/data_five_gau.rds")
training_data_gau <- read_rds("data/five_gau_clusters/data_five_gau_training.rds")
test_data_gau <- read_rds("data/five_gau_clusters/data_five_gau_test.rds")

@@

<!--add trimesh for all five-->
<!--add three screenshots from each technique with youtube links-->

>>>>

## Import data
training_data_gau <- read_rds("data/five_gau_clusters/data_five_gau_training.rds")

@@

To investigate which is the reasonable representation to visualize the five spherical Gaussian cluster data or all NLDR methods provide equivalent information, we visualize all the models in \pD{} space. Models from tSNE, UMAP, and PHATE show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, and @fig-gau3_sc), while PaCMAP and TriMAP have a long edge that connects neighboring two clusters and three well-separated clusters (see @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure.

Models from all NLDR methods show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE, UMAP, and PHATE display clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1_sc, @fig-gau2_sc, and @fig-gau3_sc). On the other hand, both TriMAP and PaCMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau4_sc and @fig-gau5_sc). Therefore, TriMAP and PaCMAP do not capture the local structure as effectively as other methods.

When compare the NLDR representations and generates models, tSNE with perplexity $61$ is the reasonable representation to visualize the five Gaussian cluster dataset.

```{r}
sc_xlims_gau <- c(-0.25, 1.25)
sc_ylims_gau <- c(-0.2, 1.2)
```

<!--trimesh with tsne-->
```{r}
gau1_scaled_obj <- gen_scaled_data(
  data = tsne_data_gau)
tsne_gau_scaled <- gau1_scaled_obj$scaled_nldr

## Compute hexbin parameters
num_bins_x_gau1 <- 12
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_gau1 <- gau1_model$df_bin_centroids
df_bin_gau1 <- gau1_model$df_bin

## Triangulate bin centroids
tr1_object_gau1 <- tri_bin_centroids(
  df_bin_centroids_gau1, x = "c_x", y = "c_y")
tr_from_to_df_gau1 <- gen_edges(
  tri_object = tr1_object_gau1)

## Compute 2D distances
distance_gau1 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_gau1,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_gau1 <- find_lg_benchmark(
  distance_edges = distance_gau1,
  distance_col = "distance")

trimesh_removed_gau1 <- vis_rmlg_mesh(
  distance_edges = distance_gau1, 
  benchmark_value = benchmark_gau1, 
  tr_coord_df = tr_from_to_df_gau1, 
  distance_col = "distance") +
  xlim(sc_xlims_gau) + ylim(sc_ylims_gau) +
  interior_annotation("a", sc_ltr_pos) 
```

<!--trimesh with umap-->
```{r}
gau2_scaled_obj <- gen_scaled_data(
  data = umap_data_gau)
umap_gau_scaled <- gau2_scaled_obj$scaled_nldr

## Compute hexbin parameters
num_bins_x_gau2 <- 44
lim1 <- gau2_scaled_obj$lim1
lim2 <- gau2_scaled_obj$lim2
r2_gau2 <- diff(lim2)/diff(lim1)

gau2_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = umap_gau_scaled,
  bin1 = num_bins_x_gau2,
  r2 = r2_gau2,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_gau2 <- gau2_model$df_bin_centroids
df_bin_gau2 <- gau2_model$df_bin

## Triangulate bin centroids
tr1_object_gau2 <- tri_bin_centroids(
  df_bin_centroids_gau2, x = "c_x", y = "c_y")
tr_from_to_df_gau2 <- gen_edges(
  tri_object = tr1_object_gau2)

## Compute 2D distances
distance_gau2 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_gau2,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_gau2 <- find_lg_benchmark(
  distance_edges = distance_gau2,
  distance_col = "distance")

trimesh_removed_gau2 <- vis_rmlg_mesh(
  distance_edges = distance_gau2, 
  benchmark_value = benchmark_gau2, 
  tr_coord_df = tr_from_to_df_gau2, 
  distance_col = "distance") +
  xlim(sc_xlims_gau) + ylim(sc_ylims_gau) +
  interior_annotation("b", sc_ltr_pos) 
```

<!--trimesh with phate-->
```{r}
gau3_scaled_obj <- gen_scaled_data(
  data = phate_data_gau)
phate_gau_scaled <- gau3_scaled_obj$scaled_nldr

## Compute hexbin parameters
num_bins_x_gau3 <- 250
lim1 <- gau3_scaled_obj$lim1
lim2 <- gau3_scaled_obj$lim2
r2_gau3 <- diff(lim2)/diff(lim1)

gau3_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = phate_gau_scaled,
  bin1 = num_bins_x_gau3,
  r2 = r2_gau3,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_gau3 <- gau3_model$df_bin_centroids
df_bin_gau3 <- gau3_model$df_bin

## Triangulate bin centroids
tr1_object_gau3 <- tri_bin_centroids(
  df_bin_centroids_gau3, x = "c_x", y = "c_y")
tr_from_to_df_gau3 <- gen_edges(
  tri_object = tr1_object_gau3)

## Compute 2D distances
distance_gau3 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_gau3,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_gau3 <- find_lg_benchmark(
  distance_edges = distance_gau3,
  distance_col = "distance")

trimesh_removed_gau3 <- vis_rmlg_mesh(
  distance_edges = distance_gau3, 
  benchmark_value = benchmark_gau3, 
  tr_coord_df = tr_from_to_df_gau3, 
  distance_col = "distance") +
  xlim(sc_xlims_gau) + ylim(sc_ylims_gau) +
  interior_annotation("c", sc_ltr_pos) 
```

<!--trimesh with pacmap-->
```{r}
gau4_scaled_obj <- gen_scaled_data(
  data = umap_data_gau)
pacmap_gau_scaled <- gau4_scaled_obj$scaled_nldr

## Compute hexbin parameters
num_bins_x_gau4 <- 18
lim1 <- gau4_scaled_obj$lim1
lim2 <- gau4_scaled_obj$lim2
r2_gau4 <- diff(lim2)/diff(lim1)

gau4_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = pacmap_gau_scaled,
  bin1 = num_bins_x_gau4,
  r2 = r2_gau4,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_gau4 <- gau4_model$df_bin_centroids
df_bin_gau4 <- gau4_model$df_bin

## Triangulate bin centroids
tr1_object_gau4 <- tri_bin_centroids(
  df_bin_centroids_gau4, x = "c_x", y = "c_y")
tr_from_to_df_gau4 <- gen_edges(
  tri_object = tr1_object_gau4)

## Compute 2D distances
distance_gau4 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_gau4,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_gau4 <- find_lg_benchmark(
  distance_edges = distance_gau4,
  distance_col = "distance")

trimesh_removed_gau4 <- vis_rmlg_mesh(
  distance_edges = distance_gau4, 
  benchmark_value = benchmark_gau4, 
  tr_coord_df = tr_from_to_df_gau4, 
  distance_col = "distance") +
  xlim(sc_xlims_gau) + ylim(sc_ylims_gau) +
  interior_annotation("d", sc_ltr_pos) 
```

<!--trimesh with trimap-->
```{r}
gau5_scaled_obj <- gen_scaled_data(
  data = trimap_data_gau)
trimap_gau_scaled <- gau5_scaled_obj$scaled_nldr

## Compute hexbin parameters
num_bins_x_gau5 <- 25
lim1 <- gau5_scaled_obj$lim1
lim2 <- gau5_scaled_obj$lim2
r2_gau5 <- diff(lim2)/diff(lim1)

gau5_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = trimap_gau_scaled,
  bin1 = num_bins_x_gau5,
  r2 = r2_gau5,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_gau5 <- gau5_model$df_bin_centroids
df_bin_gau5 <- gau5_model$df_bin

## Triangulate bin centroids
tr1_object_gau5 <- tri_bin_centroids(
  df_bin_centroids_gau5, x = "c_x", y = "c_y")
tr_from_to_df_gau5 <- gen_edges(
  tri_object = tr1_object_gau5)

## Compute 2D distances
distance_gau5 <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_gau5,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_gau5 <- find_lg_benchmark(
  distance_edges = distance_gau5,
  distance_col = "distance")

trimesh_removed_gau5 <- vis_rmlg_mesh(
  distance_edges = distance_gau5, 
  benchmark_value = benchmark_gau5, 
  tr_coord_df = tr_from_to_df_gau5, 
  distance_col = "distance") +
  xlim(sc_xlims_gau) + ylim(sc_ylims_gau) +
  interior_annotation("e", sc_ltr_pos) 
```

<!--add trimesh for all five-->

```{r}
#| label: fig-trimesh-gau
#| echo: false
#| fig-cap: "Model generated with five diiferent NLDR methods in $2\\text{-}D$ with approximately $65$ non-empty bins."
#| fig-width: 8
#| fig-height: 2
#| out-width: 100%
#| fig-pos: H

trimesh_removed_gau1 + trimesh_removed_gau2 +
  trimesh_removed_gau3 + trimesh_removed_gau4 + trimesh_removed_gau5 +
  plot_layout(ncol = 5)
```

<!--add three screenshots from each technique with you tube links-->

### 2024-05-20 paper.tex
Add draft for simulated data

>>>>

To investigate which is the reasonable representation to visualize the
five spherical Gaussian cluster data or all NLDR methods provide
equivalent information, we visualize all the models in \pD{} space.
Models from tSNE, UMAP, and PHATE show five well-separated clusters (see
\textbf{?@fig-gau1\_sc}, \textbf{?@fig-gau2\_sc}, and
\textbf{?@fig-gau3\_sc}), while PaCMAP and TriMAP have a long edge that
connects neighboring two clusters and three well-separated clusters (see
\textbf{?@fig-gau4\_sc}, and \textbf{?@fig-gau5\_sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure.

Models from all NLDR methods show five well-separated clusters (see
\textbf{?@fig-gau1\_sc}, \textbf{?@fig-gau2\_sc},
\textbf{?@fig-gau3\_sc}, \textbf{?@fig-gau4\_sc}, and
\textbf{?@fig-gau5\_sc}). This suggests that for the five Gaussian
cluster dataset, all NLDR methods effectively preserve the global
structure. tSNE, UMAP, and PHATE display clusters with varying
densities, indicating their ability to capture within-cluster variation
(see \textbf{?@fig-gau1\_sc}, \textbf{?@fig-gau2\_sc}, and
\textbf{?@fig-gau3\_sc}). On the other hand, both TriMAP and PaCMAP show
clusters with flat surfaces, suggesting a failure to capture
within-cluster variation (see \textbf{?@fig-gau4\_sc} and
\textbf{?@fig-gau5\_sc}). Therefore, TriMAP and PaCMAP do not capture
the local structure as effectively as other methods.

When compare the NLDR representations and generates models, tSNE with
perplexity \(61\) is the reasonable representation to visualize the five
Gaussian cluster dataset.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-trimesh-gau-1.pdf}

\caption{\label{fig-trimesh-gau}Model generated with five diiferent NLDR
methods in \(2\text{-}D\) with approximately \(65\) non-empty bins.}
